\documentclass{beamer}
%
% Choose how your presentation looks.
%
% For more themes, color themes and font themes, see:
% http://deic.uab.es/~iblanes/beamer_gallery/index_by_theme.html
%
\mode<presentation>
{
  \usetheme{default}      % or try Darmstadt, Madrid, Warsaw, ...
  \usecolortheme{default} % or try albatross, beaver, crane, ...
  \usefonttheme{default}  % or try serif, structurebold, ...
  \setbeamertemplate{navigation symbols}{}
  \setbeamertemplate{caption}[numbered]
  \setbeamertemplate{footline}[frame number]
  \setbeamertemplate{itemize items}[circle]
  \setbeamertemplate{theorems}[numbered]
  \setbeamercolor*{structure}{bg=white,fg=blue}
  \setbeamerfont{block title}{size=\normalsize}
}

% \newtheorem{proposition}[theorem]{Proposition}
% \theoremstyle{definition}
% \newtheorem{algorithm}[theorem]{Algorithm}
% \newtheorem{idea}[theorem]{Idea}

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{aligned-overset}
\usepackage{alltt}
\usepackage{amsmath}
\usepackage{csquotes}
% \usepackage{multicol}
% \usepackage{stmaryrd}
\usepackage{tabularx}

% \renewcommand\tabularxcolumn[1]{m{#1}}
% \newcolumntype{R}{>{\raggedleft\arraybackslash}X}=

\newcommand\logeq{\mathrel{\vcentcolon\Leftrightarrow}}
\def\code#1{\texttt{\frenchspacing#1}}
\def\padding{\vspace{0.5cm}}
\def\spadding{\vspace{0.25cm}}
\def\b{\textcolor{blue}}
\def\r{\textcolor{red}}
\def\g#1{{\usebeamercolor[fg]{block title example}{#1}}}

% fix for \pause in align
\makeatletter
\let\save@measuring@true\measuring@true
\def\measuring@true{%
  \save@measuring@true
  \def\beamer@sortzero##1{\beamer@ifnextcharospec{\beamer@sortzeroread{##1}}{}}%
  \def\beamer@sortzeroread##1<##2>{}%
  \def\beamer@finalnospec{}%
}
\makeatother

\title[Theoretical Computer Science]{Theoretical Computer Science \\ Context-Free Languages}
\author{Jonas Hübotter}
\date{}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Outline}
 \tableofcontents[subsubsectionstyle=hide]
\end{frame}
% \AtBeginSection[]
%   {
%      \begin{frame}[allowframebreaks]{Plan}
%      \tableofcontents[currentsection, sectionstyle=show/hide, hideothersubsections]
%      \end{frame}
%   }

\section{Overview}

\begin{frame}{Overview}
    \begin{block}{Representations of context-free languages}\pause
        \begin{itemize}
            \item Context-Free Grammar (CFG)\pause
            \item Pushdown Automaton (PDA)
        \end{itemize}
    \end{block}
\end{frame}

\section{Context-Free Grammar (CFG)}

\subsection{Variables}

\begin{frame}{Variables}
    \begin{definition}
        Given a grammar $G = (V, \Sigma, P, S)$, a variable $X \in V$ is
        \begin{itemize}
            \item \b{generative} if $\exists X \to_G^* w \in \Sigma^*$\pause;
            \item \b{reachable} if $\exists S \to_G^* X$\pause; and
            \item \b{helpful} if it is generative and reachable.
        \end{itemize}
    \end{definition}
\end{frame}

\subsection{Inductive Definition}

\begin{frame}{Inductive Definition}
    Given a context-free grammar $G = (V, \Sigma, P, S)$ with $V = \{A_1, \dots, A_k\}$,\pause\par
    productions $A_i \to w_0 A_{i_1}w_1 \dots w_{n-1} A_{i_n} w_n$\pause\par
    correspond to \begin{align*}
        &u_1 \in L_G(A_{i_1}) \land \dots \land u_n \in L_G(A_{i_n}) \\
        &\implies w_0 u_1 w_1 \dots w_{n-1} u_n w_n \in L_G(A_i).
    \end{align*}\pause
    Hence, $L(G) = L_G(S)$.\pause\par\spadding
    Productions produce words \r{top-down},\par
    inductive definition \textit{produces} words \r{bottom-up}.
\end{frame}

\subsection{Decomposition Lemma}

\begin{frame}{Decomposition Lemma}
    \begin{lemma}[Decomposition Lemma]
        Any derivation of length $n$ of $\beta$ from $\alpha_1 \alpha_2$ may \textit{split} $\beta$ into two separately derivable parts $\beta_1$ and $\beta_2$ at any position.\pause\ Formally:
        \begin{align*}
            \alpha_1 \alpha_2 \to_G^n \beta \iff& \exists \beta_1, \beta_2, n_1, n_2.\ \beta = \beta_1 \beta_2 \land n = n_1 + n_2\ \land \\
                                                & \alpha_1 \to_G^{n_1} \beta_1 \land \alpha_2 \to_G^{n_2} \beta_2.
        \end{align*}
    \end{lemma}
\end{frame}

\subsection{Syntax Tree}

\begin{frame}{Syntax Tree}
    \begin{definition}
        A \b{syntax tree} of a derivation $\to_G$ given $G = (V, \Sigma, P, S)$ is a tree where\pause
        \begin{enumerate}
            \item every leaf is labeled with a symbol in $\Sigma \cup \{\epsilon\}$\pause;
            \item every inner node is labeled with $A \in V$,\par
                assuming its children are $X_1, \dots, X_n \in V \cup \Sigma \cup \{\epsilon\}$, $A \to X_1 \dots X_n \in P$\pause; and
            \item a leaf labeled $\epsilon$ is an only child of its parent.
        \end{enumerate}\pause\spadding
        The \b{border} of a syntax tree is the labels of its leafs concatenated from left to right.
    \end{definition}\pause
    \begin{align*}
        A \to_G^* w &\iff w \in L_G(A) \\
                    &\iff \exists\ \text{syntax tree with root } A \text{ and border } w.
    \end{align*}
\end{frame}

\begin{frame}{Syntax Tree}
    \begin{definition}
        \begin{itemize}
            \item A CFG $G$ is \b{ambiguous} if $\exists w \in L(G)$ that has two distinct syntax trees.
            \item A CFL $L$ is \b{inherently ambiguous} if every CFG $G$ with $L(G) = L$ is ambiguous.
        \end{itemize}
    \end{definition}
\end{frame}

\subsection{Chomsky Normal Form}

\begin{frame}{Chomsky Normal Form}
    \begin{definition}[Chomsky Normal Form]
        All productions are of the form $A \to a$ or $A \to BC$ for $a \in Sigma$ and $A,B,C \in V$.
    \end{definition}\pause

    \begin{block}{Algorithm to convert a CFG to Chomsky Normal Form ($\mathcal{O}(|P|^2)$)}
        \begin{enumerate}
            \item replace every $a \in \Sigma$ occurring in a production with length $>1$ by a non-terminal\pause
            \item replace $A \to B_1 \dots B_k$ (where $k>2$) with $A \to B_1 C_2, C_2 \to B_2, \dots, C_k \to B_k$\pause
            \item remove $\epsilon$-productions (i.e. $A \to \epsilon$)\pause
            \item remove chain productions (i.e. $A \to B$)
        \end{enumerate}
    \end{block}
\end{frame}

\subsection{Other Normal Forms}

\begin{frame}{Other Normal Forms}
    \begin{definition}[Greibach Normal Form]
        All productions are of the form $A \to a A_1 \dots A_n$ for $a \in Sigma$ and $A_1, \dots, A_n \in V$.
    \end{definition}\pause

    \begin{definition}[Backus-Naur Normal Form]
        Allows the use of regular expressions in productions (in addition to symbols).
    \end{definition}
\end{frame}

\subsection{Cocke-Younger-Kasami Algorithm (CYK)}

\begin{frame}{Cocke-Younger-Kasami Algorithm (CYK)}
    Solves the word problem for CFGs.\pause

    \begin{block}{Algorithm ($\mathcal{O}(|w|^3)$)}
        Given $G = (V, \Sigma, P, S)$ in Chomsky normal form and $w = a_1 \dots a_n \in \Sigma^*$.\pause\par
        Define $V_{ij} = \{A \in V \mid A \to_G^* a_i \dots a_j\}$ for $i \leq j$ as the set of all initial symbols that may be used to derive $a_i \dots a_j$.\pause\par
        Then $w \in L_G(A) \iff A \in V_{1n}$.\pause\par\spadding
        Recursive definition of $V_{ij}$:
        \begin{itemize}
            \item base: $V_{ii} = \{A \in V \mid (A \to a_i) \in P\}$\pause
            \item step: \begin{align*}
                V_{ij} = \{A \in V \mid \substack{\exists i \leq k < j, B \in V_{ik}, C \in V_{(k+1)j}.\ \\ (A \to BC) \in P}\}
            \end{align*}
        \end{itemize}
    \end{block}
\end{frame}

\section{Pushdown Automaton (PDA)}

\begin{frame}{PDA}
    \begin{definition}
        A \b{pushdown automaton (PDA)} $M = (Q, \Sigma, \Gamma, q_0, Z_0, \delta, F)$\pause\ consists of
        \begin{itemize}
            \item a finite set of \b{states} $Q$\pause;
            \item a (finite) \b{input alphabet} $\Sigma$\pause;
            \item a (finite) \b{stack alphabet} $\Gamma$\pause;
            \item an \b{initial state} $q_0 \in Q$\pause;
            \item an \b{initial stack element} $Z_0 \in \Gamma$\pause;
            \item a (partial) \b{transition function} $\delta: Q \times (\Sigma \cup \{\epsilon\}) \times \Gamma \to 2^{Q \times \Gamma^*}$\pause; and
            \item a set of \b{terminal (accepting) states} $F \subseteq Q$.
        \end{itemize}
    \end{definition}\pause
    Graphically, transitions are denoted as $a, Z/\alpha$ where $a \in \Sigma$ is the input, $Z \in \Gamma$ is the top stack element, and $\alpha \in \Gamma^*$ replaces $Z$ in the new stack.
\end{frame}

\begin{frame}{PDA}
    \begin{definition}
        The \b{configuration} of a PDA $M$ is a triple $(q, w, \alpha)$ where $q \in Q$ is its state, $w \in \Sigma^*$ is its remaining input, and $\alpha \in \Gamma^*$ is its stack.\pause\par\spadding
        The \b{initial configuration} of $M$ on input $w \in \Sigma^*$ is $(q_0,w,Z_0)$.
    \end{definition}\pause
    \begin{definition}
        The \b{transition relation} of a PDA $M$ is \begin{align*}
            (q,aw,Z\alpha) \to_M (q',w,\beta\alpha) &\quad\text{if } (q',\beta) \in \delta(q,a,Z) \\
            (q,w,Z\alpha) \to_M (q',w,\beta\alpha)  &\quad\text{if } (q',\beta) \in \delta(q,\epsilon,Z).
        \end{align*}
    \end{definition}
\end{frame}

\begin{frame}{PDA}
    \begin{definition}
        PDA $M$ \b{accepts} $w \in \Sigma^*$ with final state if \begin{align*}
            (q_0,w,Z_0) \to_M^* (f,\epsilon,\gamma) \quad\text{for } f \in F, \gamma in \Gamma^*.
        \end{align*}
        So, $L_F(M) = \{w \in \Sigma^* \mid \exists f \in F, \gamma \in \Gamma^*.\ (q_0,w,Z_0) \to_M^* (f,\epsilon,\gamma)\}$.
    \end{definition}\pause
    \begin{definition}
        PDA $M$ \b{accepts} $w \in \Sigma^*$ with empty stack if \begin{align*}
            (q_0,w,Z_0) \to_M^* (q,\epsilon,\epsilon) \quad\text{for } q \in Q.
        \end{align*}
        So, $L_{\epsilon}(M) = \{w \in \Sigma^* \mid \exists q \in Q.\ (q_0,w,Z_0) \to_M^* (q,\epsilon,\epsilon)\}$.
    \end{definition}\pause\padding
    \r{Both accepting conditions are equally powerful.}
\end{frame}

\subsection{Lemmas}

\begin{frame}{Lemmas}
    \begin{lemma}[Extension Lemma]
        Every derivation may occur as a sub-derivation of a larger derivation\pause:
        \begin{align*}
            (q,u,\alpha) \to_M^n (q',u',\alpha') \implies (q,uv,\alpha\beta) \to_M^n (q',u'v,\alpha'\beta).
        \end{align*}
    \end{lemma}\pause
    \begin{lemma}[Decomposition Lemma]
        Every derivation that empties the stack can be divided into sub-derivations that each remove a single symbol from the stack\pause:
        Given $(q,w,Z_1 \dots Z_k) \to_M^n (q',\epsilon,\epsilon)$\pause,\par
        then $\forall i \in [1,k].\ \exists u_i, p_i, n_i$ such that \begin{align*}
            (p_{i-1},u_i,Z_i) \to_M^{n_i} (p_i,\epsilon,\epsilon)
        \end{align*} with $w = u_1 \dots u_k$, $q = p_0$, $q_k = p_k$, and $n = \sum_{i=1}^k n_i$.
    \end{lemma}
\end{frame}

\subsection{CFG $\to$ PDA}

\begin{frame}{CFG $\to$ PDA}
    Given CFG $G = (V,\Sigma,P,S)$,\pause
    \begin{enumerate}
        \item bring all productions into the form \begin{align*}
            A \to b B_1 \dots B_k \quad\text{for } b \in \Sigma \cup \{\epsilon\}
        \end{align*}\pause
        \item define the PDA $M = (\{q\},\Sigma,V,q,S,\delta)$\pause\ with \begin{align*}
            \delta(q,b,A) = \{(q,\beta) \mid (A \to b \beta) \in P\}.
        \end{align*}
    \end{enumerate}\pause
    Then, $L(G) = L_{\epsilon}(M)$.
\end{frame}

\subsection{PDA $\to$ CFG}

\begin{frame}{PDA $\to$ CFG}
    Given PDA $G = (Q,\Sigma,\Gamma,q_0,Z_0,\delta,F)$,\pause\ define CFG $G = (V,\Sigma,P,S)$.\pause\par\spadding
    We define $V = Q \times \Gamma \times Q \cup \{S\}$ where each $[q,Z,p] \in V$ describes all possibilities of going from state $q \in Q$ to state $p \in Q$ while $Z \in \Gamma$ is the top stack element.\pause\par\spadding
    We define the productions $P$ as \begin{itemize}
        \item $\forall q \in Q.\ S \to [q_0,Z_0,q]$\pause\ and
        \item $\forall (r_0, Z_1 \dots Z_k) \in \delta(q,b,Z).\ \forall r_1, \dots, r_k \in Q.$\pause \begin{align*}
            [q,Z,r_k] \to b [r_0,Z_1,r_1][r_1,Z_2,r_2]\dots[r_{k-1},Z_k,r_k].
        \end{align*}
    \end{itemize}\pause\padding
    We observe that \begin{align*}
        [q,Z,r_k] \to_G^* w \iff (q,w,Z) \to_M^* (r_k,\epsilon,\epsilon).
    \end{align*}\pause
    So, $L(G) = L_{\epsilon}(M)$.
\end{frame}

\subsection{Deterministic Pushdown Automaton (DPDA)}

\section{Closure Properties}

\begin{frame}{Closure Properties}
    \begin{theorem}
        Given the context-free languages $L, L_1, L_2$, then the following are also centext-free languages:\pause
        \begin{itemize}
            \item $L_1 L_2$\pause;
            \item $L_1 \cup L_2$\pause; and
            \item $L^*$.
        \end{itemize}
    \end{theorem}\pause

    \begin{theorem}
        Given the deterministic context-free language $L$, then $\bar{L}$ is deterministic context-free.
    \end{theorem}
\end{frame}

\section{Pumping Lemma}

\begin{frame}{Pumping Lemma}
    \begin{lemma}[Pumping Lemma for context-free languages]
        Let $L \subseteq \Sigma^*$ be context-free.\pause\ Then there exists some $n > 0$ such that every $z \in L$ with $|z| \geq n$ can be decomposed into $z = uvwxy$\pause\ such that
        \begin{itemize}
            \item $vx \neq \epsilon$\pause;
            \item $|vwx| \leq n$\pause; and
            \item $\forall i \geq 0.\ uv^iwx^iy \in L$.
        \end{itemize}
    \end{lemma}\pause
    \r{A necessary condition for context-free languages.}
\end{frame}

\begin{frame}{Pumping Lemma}
    \begin{example}[proof structure]
        Assume $L$ is context-free.\par
        Let $n > 0$ be a Pumping Lemma number.\pause\par
        Choose $z \in L$ with $|z| \geq n$.\par
        Define $z = uvwxy$ with $vx \neq \epsilon$ and $|vwx| \leq n$.\pause\par
        Then, $\forall i \geq 0.\ uv^iwx^iy \in L$.\pause\par
        Now, use the last statement to find a contradiction separating all possible cases for $v$ and $x$.
    \end{example}
\end{frame}

\end{document}
